---
title: "Finding Important Covariates"
output: html_notebook
---

```{r}
library(dplyr)       # data wrangling
library(rpart)       # performing regression trees
library(rpart.plot)  # plotting regression trees
library(countrycode)
library(data.table)
library(ggplot2)
```


```{r}
data<-read.csv("data/Data for R.csv", na.strings = "")
```

TODO: decide on handling of outliers
```{r}
#data <- subset(data, Outlier_NT ==0 | Outlier_CT == 0)
```

# Preliminary Exploration of Covariates and Treatment Vars

While almost all covariates are constant for a pair, there are a few "covariates" that differ within the pair, all of which are differening management practices, and which thus are more appropriate to be treated as an additional treatment variable. Before defining our covariates and treatments, we first check whether these actually do differ between pairs. 
```{r}
potential_mgmt_practice_treatments <- c("R.CT",
                          "R.NT", 
                          "SC.CT",  
                          "SC.NT",
                          "Rotation.CT",
                          "Rotation.NT",
                          "Residue.Management.in.CT.in.last.cropping.season..available.in.current.cropping.season.",
                          "Residue.Management.in.NT.in.last.cropping.season..available.in.current.cropping.season.", 
                          "Weed.and.pest.control.CT",
                          "Weed.and.pest.control.NT", 
                          "Fertilization.CT", 
                          "Fertilization.NT", 
                          "Irrigation.CT", 
                          "Irrigation.NT"
                          ) 

for (i in seq(1,length(potential_mgmt_practice_treatments),2)){
  all_same <- all(data[,potential_mgmt_practice_treatments[i]] == data[,potential_mgmt_practice_treatments[i+1]], na.rm = TRUE)
  if (all_same) {
    print(potential_mgmt_practice_treatments[i])
  }

}

```

We see that R (Crop Rotation w/ more than 2 conservation crops) and Irrigation are always the same across pairs, so we treat these as covariates, rather than treatments, dropping the duplicated column, and renaming to reflect this change. 
```{r}
 data_clean <- data %>% 
  rename(Rotation.Conservation.Crop = R.CT, Irrigation = Irrigation.CT)
```

We also choose not to include any columns that have more than 50% of values missing. (Of these, only  Cover.crop.before.sowing and Irrigation would have been variables of interest).
```{r}

perc_null <- sapply(X = data_clean, FUN = function(x) sum(is.na(x))/length(x))
print(perc_null[perc_null > .5])
```



# Defining Covariates of Interest
There are various covariates (as well as other additional treatments) of interest that fall into different categories. 
Note that Study id (Seq) is not of interest, but is needed to later identify undesired stratifications containing observations from only a small number of studies. 
```{r}

study_covars <- c('Seq') #id for study 
study_location_covars <- c( "Site.country", 
                            "Latitude",
                            "Longitude")

study_year_covars <- c("Start.year.of.NT...or.first.year.of.experiment.if.missing.", 
                     "Sowing.year",
                     "Harvest.year")

seasonal_covars <- c("Sowing.month", 
                     "Harvesting.month"
                    )

mgmt_practice_treatments <- c("SC.CT",  #categorical: Yes, No or Mixed
                          "SC.NT",
                          "Rotation.CT",# indicator of any crop rotation
                          "Rotation.NT",
                          "Residue.Management.in.CT.in.last.cropping.season..available.in.current.cropping.season.",# categorical: retained, removed or average
                          "Residue.Management.in.NT.in.last.cropping.season..available.in.current.cropping.season.",
                          "Weed.and.pest.control.CT", #indicator
                          "Weed.and.pest.control.NT",
                          "Fertilization.CT", #categorical: Yes, No or Mix
                          "Fertilization.NT",
                          "Durationof.NT.period.at.sowing..yrs." #duration of treatment (in years)
                          ) 


mgmt_practice_covars <- c("Rotation.Conservation.Crop", #indicator of Crop rotation with >= 3 crops in accordance with FAO def of conservation agriculture 
                          "N.input" #Yes, No or Mix
                          ) 

climatic_soil_vars <- c("pH..surface.layer.", #at experiment site at time of experiment start
                        "P", #Precipitation over the growing season	(mm)
                        "E", #Potential evapotranspiration over the growing season (mm))
                        "PB", #Precipitation balance (mm) := the amount of available water in the growing season for rainfed field
                        "Tave", #Average air temperature during the growing season (C)
                        "Tmax", #Max air temperature during the growing season (C) 
                        "Tmin", #Min air temperature during the growing season (C) 
                        "ST") #Soil texture at experiment  location. Categorical: Sandy Loam; Loam; Silt Loam; Sandy Clay Loam; Clay Loam; Sandy Clay; Clay
                        
crop_covar<- c("Crop")
outcome <- c("Ln_ratio")

keep_cols <- c(study_covars,study_location_covars, study_year_covars, seasonal_covars,crop_covar, mgmt_practice_covars, mgmt_practice_treatments, climatic_soil_vars, outcome)
```


```{r}
data_clean<- data_clean[,keep_cols]
```

In 2 cases, experimement year is before sowing year, which is likely a data entry error, so we remove these observation pairs:
```{r}
data_clean <- data_clean[data_clean$Durationof.NT.period.at.sowing..yrs. >= 0,]
```


The ph covariate is currently coded as a string and we need to recode this as numeric, handling cases where string contains a range or other edge cases. 
```{r}
#set ph vars for 22 cases where different for treatment and control to Null (also the one case with a comma, "5,2" since unclear what this means-- likely a typo)
msk<- grepl("CT|NT|,", data_clean$pH..surface.layer.)
print(sum(msk))
data_clean[msk,]$pH..surface.layer. <-NA
#remove "with lime application" from end of string in 15 cases
msk<- grepl(" with lime application", data_clean$pH..surface.layer.)
print(sum(msk))
data_clean[msk,]$pH..surface.layer. <- substr(data_clean[msk,]$pH..surface.layer., 1,10)
#remove "or above" from 10 cases where ph var is "\d or above" (treat as \d)
msk<- grepl("above", data_clean$pH..surface.layer.)
print(sum(msk))
data_clean[msk,]$pH..surface.layer. <- gsub("([0-9]+).*", "\\1", data_clean[msk,]$pH..surface.layer.)
# for 164 cases that give a range with "to" or "-": replace "{x1} to/- {x2}" strings with mean of x1 and x2
msk<- grepl("to", data_clean$pH..surface.layer.)
print(sum(msk))
range_strings <- data_clean[msk,]$pH..surface.layer. 
data_clean[msk,]$pH..surface.layer. <- sapply(strsplit(range_strings, split = " to ", fixed = TRUE), function(k) mean(as.numeric(k)))
msk<- grepl("-", data_clean$pH..surface.layer.)
print(sum(msk))
range_strings <- data_clean[msk,]$pH..surface.layer. 
data_clean[msk,]$pH..surface.layer. <- sapply(strsplit(range_strings, split = "-", fixed = TRUE), function(k) mean(as.numeric(k)))
```



Convert character columns to factors and and cleaned pd column to numeric
```{r}
to_factor_cols <- c(c("Seq", "Site.country", "Crop", 'ST'), mgmt_practice_covars)
data_clean[to_factor_cols] <- lapply(data_clean[to_factor_cols], factor)
data_clean$pH..surface.layer. <- as.numeric(data_clean$pH..surface.layer.)

data_covar <- data_clean[, !names(data_clean) %in% mgmt_practice_treatments]
```

For each of our management practice treatment pairs there are 2 levels, 1=yes and 0=no, and sometimes a third level "Mixed" (For residue management the levels are 1=Retained, 0=Removed, and Average. Note that these are sometimes null. We focus on just the binary levels, and create a factor of four levels corresponding to each possible pair $Y(CT,0), Y(CT,1), Y(NT,0), Y(NT,1)$, where not all levels are always in the data. 

```{r}
make_treatment_pairs <- function(df, c_treatment_col, t_treatment_col, new_col,  binary_treatments = c("Yes", "No")) {
  new_col_name <- quo_name(new_col)
  df_new <- df %>%
    mutate(!!new_col_name := ifelse(.data[[c_treatment_col]] %in% binary_treatments & .data[[t_treatment_col]] %in% binary_treatments & .data[[c_treatment_col]] == .data[[t_treatment_col]],
                         paste(.data[[c_treatment_col]], .data[[t_treatment_col]]),
                         NA))
  
  df_new <- df_new[!names(df_new) %in% c(c_treatment_col, t_treatment_col)]
  return (df_new)
}
```

```{r}
data_clean <- make_treatment_pairs(df = data_clean, 
                                c_treatment = "SC.CT",
                                t_treatment = "SC.NT",
                                new_col = 'Soil.Cover.Pair')

data_clean <- make_treatment_pairs(df = data_clean, 
                                c_treatment = "Rotation.CT",
                                t_treatment = "Rotation.NT",
                                new_col = 'Crop.Rotation.Pair')

data_clean <- make_treatment_pairs(df = data_clean, 
                                c_treatment = "Residue.Management.in.CT.in.last.cropping.season..available.in.current.cropping.season.",
                                t_treatment = "Residue.Management.in.NT.in.last.cropping.season..available.in.current.cropping.season.",
                                new_col = 'Residue.Management.Pair',
                                binary_treatments = c("retained", "removed"))

data_clean <- make_treatment_pairs(df = data_clean, 
                                c_treatment = "Weed.and.pest.control.CT",
                                t_treatment = "Weed.and.pest.control.NT",
                                new_col = 'Weed.Pest.Control.Pair')

data_clean <- make_treatment_pairs(df = data_clean, 
                                c_treatment = "Fertilization.CT",
                                t_treatment = "Fertilization.NT",
                                new_col = 'Fertilization.Pair')
```


We now consider duration of treatment as treatment levels
```{r}
hist(data_clean$Durationof.NT.period.at.sowing..yrs., breaks = 52)
```


```{r}
dur_max <- max(data_clean$Durationof.NT.period.at.sowing..yrs.)
data_clean<-data_clean %>%
  mutate(treatment_duration_4_grouping = cut(Durationof.NT.period.at.sowing..yrs., 
                                  breaks = c(0,5, 10,20,dur_max), 
                                  include.lowest = TRUE),
         treatment_duration_eq_3 = cut_number(Durationof.NT.period.at.sowing..yrs.,3), 
         treatment_duration_eq_4 = cut_number(Durationof.NT.period.at.sowing..yrs.,4), 
         treatment_duration_eq_5 = cut_number(Durationof.NT.period.at.sowing..yrs.,5))

data_clean <- data_clean[!names(data_clean) %in% c("Durationof.NT.period.at.sowing..yrs.")]

```


We now have a dataframe with the covariates, and one with the various treatments.
```{r}

mgmt_practice_treatment_pairs <- c('Soil.Cover.Pair', 'Crop.Rotation.Pair','Residue.Management.Pair', 'Weed.Pest.Control.Pair', 'Fertilization.Pair')
treatment_durs <- c('treatment_duration_4_grouping', 
                    'treatment_duration_eq_3',
                    'treatment_duration_eq_4', 
                    'treatment_duration_eq_5')

data_covar <- data_clean[, !names(data_clean) %in% c(mgmt_practice_treatment_pairs,treatment_durs)]

data_additional_treatment <- data_clean[, c(mgmt_practice_treatment_pairs, treatment_durs, outcome, crop_covar)]
```
